Testing traversability of protein tunnels with flexible ligands using motion planning

### 2017-06-13 paper.tex
images of m003
alg updated

\def\C{\mathcal{C}}
\def\CF{\mathcal{C}_{free}}
\def\CFD{{\mathcal{C}^s_{free}}}
\def\dt{d_{goal}}
\def\da{d_{atom}}

@@

%spacing for algorithm environment. 1.0 mean normal spacing
\def\gb{p_{c}}

@@

Proteins are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
%We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
%   can be detected as atoms located at $\alpha$-shape of the molecule.
%The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
A tunnel is described by a sequence of collision-free spheres $T=( (t_i, r_i) )$, where $n$ denotes the number of spheres,
$t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $t_i$. 
The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.

@@

Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
To decrease the computational burden, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
%For each conformation $c\in \L$ an energy potential is also calculated.
%Let $\L_i$ denote the set of conformations that 
We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.
%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
%can be switched to the confomration $c_j \in C$.

The proteins are dense structures with narrow tunnels.
Depending on the protein, the tunnel bottlenecks can be even less then $1$~\AA, which is too narrow for ligands with more than 2 atoms.
For the purpose of traversability study, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.

A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., it's geometric center), rotation around $x$, $y$ and $z$ axes,
index of the conformation $l\in \L$ and the actual scale $s$.
A discrete set of scales is used, i.e., $s \in {\smin, \smin+\sdelta, \ldots, \smax}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
All possible configurations form the configuration space $\C$. 
A configuration is collision-free if there is no intersection between the protein atoms and ligand atoms scaled by $s$.
The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

@@

The task of the traversability study is to determine if a given ligand can traverse  a given tunnel in the protein.
The tunnels are usually computed on proteins without ligands, which implies narrow void space.

The configuration space is searched using the RRT principle~\cite{lavalleRRT}.
The main loop of the method is described in Alg.~\ref{alg::main}.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between the $\qrand$ and the tree is realized using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
is guided using moving virtual goal~\cite{vonasek2009rrt}.
The virtual goal $v$ is the index of the actual sphere of the tunnel, around which the samples
are generated with probability $\gb$. 
After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
The algorithm terminates after a predefined planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = N$.

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::main}main loop of the RRT planner}
\KwIn{
    tunnel $T=(t_i, r_i), n=1,\ldots N$ represented by $N$ spheres,
    initial configuration $\qinit$,
    goal configuration $\qgoal$,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 0$; // index of the virtual goal\\
$iteration = 0$\;
\While{$iteration < \Imax$ and $v < N$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around actual virtual goal $t_n \in T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;

@@

    \For{$i=N : v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree to $P[i]$\;
        \If{$d < DT$}{
            $v = i$; // new virtual goal found\;
            break\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
\end{algorithm}

@@

>>>>

\def\C{\mathcal{C}}
\def\CF{\mathcal{C}_{free}}
\def\CFD{{\mathcal{C}^s_{free}}}
\def\dt{d_{tunnel}}
\def\da{d_{atom}}

@@

%spacing for algorithm environment. 1.0 mean normal spacing
\def\gb{p_{tunnel}}

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
%Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
%We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
%   can be detected as atoms located at $\alpha$-shape of the molecule.
%The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
The flexiblity of a ligand is modeled using the set $\L$ of its conformations.
The conformations can be prepared e.g., considering the potential energy using tools like Rosetta.
We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.

%Another approach is to model the ligand flexibility using a predefined set of feasible conformations $\L$
%To decrease the computational burden, a set of feasible conformations $\L$ is prepared, e.g. using Rosseta tool.
%The conformations can be prepared using Molecular Dynamics software such as Rosetta.
%For each conformation $c\in \L$ an energy potential is also calculated.
%Let $\L_i$ denote the set of conformations that 
%The transition matrix $M$ defines the possible changes of conformations, i.e., $M(i,j)=1$ if the conformation $c_i \in C$
%can be switched to the confomration $c_j \in C$.

A protein tunnel is described by a sequence of collision-free spheres $T=( (t_i, r_i) )$, $i=1,\ldots,n$, where $n$ denotes the number of spheres,
$t_i=(x_i,y_i,z_i)$ is their 3D position and $r_i$ denotes the maximum collision-free radius of a sphere centered at $t_i$. 
The tunnels can be found by tunnel detection tools like CAVER 3.0~\cite{caver3}.

@@

The proteins are dense structures with narrow tunnels.
Depending on the protein, the tunnel bottlenecks can be even less then $1$~\AA, which is too narrow for ligands with more than 2 atoms.
For the purpose of traversability study, the atomic radii of the ligand are scaled down by a factor $s, 0 < s \le 1$.
A discrete set of scales is used, i.e., $s \in \{\smin, \smin+\sdelta, \ldots, \smax\}$, where 
$\smin$ is the minimal allowed scale, $\smax=1$ is the maximal allowed scale and $\sdelta$ is the minimal difference between two scales.
This scaling technique is used also in other tools like MoMaLigPath~\cite{cortes2005path}.

A configuration of the pseudoflexible ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$  is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g., it's geometric center), rotation around $x$, $y$ and $z$ axes,
index of the conformation $l\in \L$ and the actual scale $s$.
All possible configurations form the configuration space $\C$. 
A configuration is collision-free if there is no intersection between the protein atoms and ligand atoms scaled by $s$.
%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

@@

The task of the traversability study is to determine if a given ligand can traverse  a given tunnel in the protein.
We propose a RRT-based method to compute the trajectories of the ligand.
As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
suffer from the well known narrow passage problem~\cite{hannaWIS}.

To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
Instead, they are generated using the spheres of the tunnel.
The main loop of the method is described in Alg.~\ref{alg::main}.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between the $\qrand$ and the tree is realized using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

To guide thee growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v$ is the index of a sphere, around which the random samples are generated with probability $\gb$.
%The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
%is guided using moving virtual goal~\cite{vonasek2009rrt}.
After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to ``skip'' such parts of the tunnels that are
not traversible or reachable by the ligand.
The algorithm terminates after a predefined planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = N$.

The random samples $\qrand$ are generated around the virtual goal.
The translation of $\qrand$ is generated from Normal distribution centered at $t_v$ with radius $R_v$.
The radius $R_v$ is set as the absolut distance from the tunnel centerline, e.g. $R_v=3\AA$.
Alternatively, $R_v$ can be set to $R_v=r_v$ according the radius of the tunnel at the virtual goal.

\linesnumbered
\begin{algorithm}[h]
%\setstretch{\straa}
\caption{\label{alg::main}main loop of the RRT planner}
\KwIn{
    tunnel $T=(t_i, r_i), n=1,\ldots N$ represented by $N$ spheres,
    initial configuration $\qinit$,
    goal configuration $\qgoal$,
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$ and $\sdelta$
}
\KwOut{
    configuration tree $\T$\;
}
\hrule
$v = 0$; // index of the virtual goal\\
$iteration = 0$\;
\While{$iteration < \Imax$ {\bf and}  $v < N$}{
    \eIf{$rand() < \gb$}{
        $\qrand$ = random sample around actual virtual goal $t_n \in T$\;
    }{
        $\qrand$ = random sample from $\C$\;
    }
    $\qnear$ = nearest node in $\T$ towards $\qrand$\;

@@

    \For{$i= N,N-1,\ldots v+1,v$}{ \nllabel{alg::main:a}
        $d$ = nearest node in the tree to $T[i]$\;
        \If{$d < \dt $}{
            $v = i+1$; // new virtual goal found\;
            {\bf  break}\;
        }
    } \nllabel{alg::main:b}
    $iteration = iteration+1$\;
}
\return $\T$\;
\end{algorithm}

@@

\def\tmpa{0.16\textwidth}

\begin{figure}
\centering
\begin{tabular}{cccccc}
\includegraphics[width=\tmpa]{fig/m003-000} &
\includegraphics[width=\tmpa]{fig/m003-001} &
\includegraphics[width=\tmpa]{fig/m003-002} & 
\includegraphics[width=\tmpa]{fig/m003-003} & 
\includegraphics[width=\tmpa]{fig/m003-004} & 
\includegraphics[width=\tmpa]{fig/m003-005} \\
\includegraphics[width=\tmpa]{fig/m003-006} &
\includegraphics[width=\tmpa]{fig/m003-007} &
\includegraphics[width=\tmpa]{fig/m003-008} & 
\includegraphics[width=\tmpa]{fig/m003-009} & 
\includegraphics[width=\tmpa]{fig/m003-010} & 
\includegraphics[width=\tmpa]{fig/m003-011} \\

\end{tabular}
\caption{todo and maybe delete}
\end{figure}
